---
title: "DESeq2 on PECHi-C data (read-counts-CIS) after PRC2 perturbation"
author: "Dr. Marina Nocente"
date: "2024-12-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Libraries
```{r}
library(data.table)
library(ggplot2)
library(DESeq2)

```

# File paths
```{r}
base_dir <- "~/Documents/Bioinformatics/PECHiC_PRC2_DPPA/Exploration_per_BaitID_PRC2"
working_dir <- "~/Documents/Project_writing/Paper_PEs/Data_release/PECHiC_PRC2"

read_counts_cis_file <- file.path(base_dir,"Read_counts_PerBait_CIS_table_Chinput.txt")

```

# Upload and prepare the PECHi-C data
```{r}
read_counts_cis <- fread(read_counts_cis_file)

read_counts_cis_df <- as.data.frame(read_counts_cis) # Convert to a data.frame
head(read_counts_cis_df)

## Replace all NA values in the data.frame with 0
read_counts_cis_df[is.na(read_counts_cis_df)] <- 0
sum(is.na(sum(is.na(read_counts_cis_df))))  # Should return 0
# Note: no detection and 0 reads are different

## Select the columns of interest (use merged files for PROTAC and UNC1999 to balance the sequencing depths across replicates)
read_counts_cis_df <- read_counts_cis_df[, c("baitID", "DMSO_R3", "DMSO_R4", "DMSO_R5", "PROTAC_merged-R3_R5", "PROTAC_R4", "UNC1999_merged_R3_R4", "UNC1999_R5")]

fwrite(read_counts_cis_df, file.path(working_dir, "Read_counts_per_PEbaits_chinput_.txt"), sep = "\t", quote=F, col.names=T)

```


# Create the DESeq2 dataset
For DESeq2 to function, you need two main inputs:
- A count matrix: This matrix should have rows as features (bait IDs) and columns as samples (DMSO, PROTAC, UNC1999, etc.).
- A sample information table (metadata): This table contains information about the experimental conditions or the groups each sample belongs to.

## The count Matrix:
```{r}
## Preparing the Count Matrix
# Remove the baitID column (the first column)
count_data <- read_counts_cis_df[, -1]  

# Assign the baitID as row names
rownames(count_data) <- read_counts_cis_df$baitID
head(count_data)

# Convert count_data into a matrix
count_data_matrix <- as.matrix(count_data)

```

## The metadata:
```{r}
# Create sample metadata
colData <- data.frame(
  condition = factor(c("DMSO", "DMSO", "DMSO", "PROTAC", "PROTAC", "UNC1999", "UNC1999")),
  row.names = colnames(count_data_matrix)  # Ensure the column names match the samples in the matrix
)

# Check the colData structure
head(colData, n=8)

```


# Proceed to DESeq2 analysis
```{r}
## Create DESeq2 dataset
dds <- DESeqDataSetFromMatrix(countData = count_data_matrix, 
                              colData = colData, 
                              design = ~ condition)

## Run DESeq2
dds <- DESeq(dds)
summary(dds)

```



# Extract the results: PROTAC versus DMSO

```{r}
## Extract results for the comparison: PROTAC vs DMSO
res_PROTAC_DMSO <- results(dds, contrast = c("condition", "PROTAC", "DMSO"))

# View the results
View(as.data.frame(res_PROTAC_DMSO))
dim(res_PROTAC_DMSO) # 31754     6

# Export the full "merged" results
res_PROTAC_DMSO_dt <- as.data.table(res_PROTAC_DMSO, keep.rownames = "baitID") 

fwrite(res_PROTAC_DMSO_dt, file.path(working_dir, "res_deseq2PECHiC_PROTAC_DMSO_merged_all.txt"),
       sep = "\t", quote = FALSE)

```


# Extract the results: UNC1999 versus DMSO

```{r}
## Extract results for the comparison: UNC1999 vs DMSO
res_UNC_DMSO <- results(dds, contrast = c("condition", "UNC1999", "DMSO"))

# View the results
View(as.data.frame(res_UNC_DMSO))
dim(res_UNC_DMSO) # 31754     6

# Export the full "merged" results
res_UNC_DMSO_dt <- as.data.table(res_UNC_DMSO, keep.rownames = "baitID") 

fwrite(res_UNC_DMSO_dt, file.path(working_dir, "res_deseq2PECHiC_UNC_DMSO_merged_all.txt"),
       sep = "\t", quote = FALSE)

```




