---
title: "RNAseq analysis after PRC2 perturbation (PROTAC and catalytic inhibitor)"
author: "Dr. Marina Nocente"
date: "2025-03-24"
output: html_document
---

# Libraries and path
```{r}
#######################################
## Libraries
#######################################
library(DESeq2)
library(AnnotationDbi)
library(org.Hs.eg.db)
library(EnhancedVolcano)

library(ggplot2)

#######################################
## File paths
#######################################
working_dir <- file.path("~/Documents/Bioinformatics/RNAseq_PRC2_perturbations/")

featureCounts_file_dir <- file.path("~/Documents/Bioinformatics/RNAseq_PRC2_perturbations/featureCounts/")

```


# Upload the featureCounts data
```{r Upload the featureCounts data}

## Load data, skipping the first line (header info)
counts_DMSO_R3 <- read.delim(file.path(featureCounts_file_dir, "lane8805_CTGATCGT_ATATGCGC_Library1_DMSO_Day5_Rep3_L001_GRCh38_hisat2_sorted_Counts.txt"), header=TRUE, comment.char="#", row.names=1)

counts_DMSO_R4 <- read.delim(file.path(featureCounts_file_dir, "lane8805_GAGACGAT_TAACCGGT_Library4_DMSO_Day5_Rep4_L001_GRCh38_hisat2_sorted_Counts.txt"), header=TRUE, comment.char="#", row.names=1)

counts_DMSO_R5 <- read.delim(file.path(featureCounts_file_dir, "lane8805_CGCATGAT_TCAGGCTT_Library7_DMSO_Day5_Rep5_L001_GRCh38_hisat2_sorted_Counts.txt"), header=TRUE, comment.char="#", row.names=1)

counts_PROTAC_R3 <- read.delim(file.path(featureCounts_file_dir, "lane8805_ACTCTCGA_TGGTACAG_Library2_PROTAC_Day_5_Rep_3_L001_GRCh38_hisat2_sorted_Counts.txt"), header=TRUE, comment.char="#", row.names=1)

counts_PROTAC_R4 <- read.delim(file.path(featureCounts_file_dir, "lane8805_CTTGTCGA_GAACATCG_Library5_PROTAC_Day5_Rep4_L001_GRCh38_hisat2_sorted_Counts.txt"), header=TRUE, comment.char="#", row.names=1)

counts_PROTAC_R5 <- read.delim(file.path(featureCounts_file_dir, "lane8805_ACGGAACA_GTTCTCGT_Library8_PROTAC_Day5_Rep5_L001_GRCh38_hisat2_sorted_Counts.txt"), header=TRUE, comment.char="#", row.names=1)

counts_UNC_R3 <- read.delim(file.path(featureCounts_file_dir, "lane8805_TGAGCTAG_AACCGTTC_Library3_UNC_Day5_Rep3_L001_GRCh38_hisat2_sorted_Counts.txt"), header=TRUE, comment.char="#", row.names=1)

counts_UNC_R4 <- read.delim(file.path(featureCounts_file_dir, "lane8805_TTCCAAGG_CCTTGTAG_Library6_UNC_Day5_Rep4_L001_GRCh38_hisat2_sorted_Counts.txt"), header=TRUE, comment.char="#", row.names=1)

counts_UNC_R5 <- read.delim(file.path(featureCounts_file_dir, "lane8805_CGGCTAAT_AGAACGAG_Library9_UNC_Day5_Rep5_L001_GRCh38_hisat2_sorted_Counts.txt"), header=TRUE, comment.char="#", row.names=1)


## Rename the 6th column which contains the counts
colnames(counts_DMSO_R3)[6] <- "Counts_DMSO_R3"
colnames(counts_DMSO_R4)[6] <- "Counts_DMSO_R4"
colnames(counts_DMSO_R5)[6] <- "Counts_DMSO_R5"
colnames(counts_PROTAC_R3)[6] <- "Counts_PROTAC_R3"
colnames(counts_PROTAC_R4)[6] <- "Counts_PROTAC_R4"
colnames(counts_PROTAC_R5)[6] <- "Counts_PROTAC_R5"
colnames(counts_UNC_R3)[6] <- "Counts_UNC_R3"
colnames(counts_UNC_R4)[6] <- "Counts_UNC_R4"
colnames(counts_UNC_R5)[6] <- "Counts_UNC_R5"

## Create a GeneID column 
counts_DMSO_R3$geneID <- rownames(counts_DMSO_R3)
counts_DMSO_R4$geneID <- rownames(counts_DMSO_R4)
counts_DMSO_R5$geneID <- rownames(counts_DMSO_R5)
counts_PROTAC_R3$geneID <- rownames(counts_PROTAC_R3)
counts_PROTAC_R4$geneID <- rownames(counts_PROTAC_R4)
counts_PROTAC_R5$geneID <- rownames(counts_PROTAC_R5)
counts_UNC_R3$geneID <- rownames(counts_UNC_R3)
counts_UNC_R4$geneID <- rownames(counts_UNC_R4)
counts_UNC_R5$geneID <- rownames(counts_UNC_R5)


## Merge all the counts in one data.frame
counts_total <- counts_DMSO_R3[, c("geneID", "Counts_DMSO_R3")]

# Merge other data frames one by one
counts_total <- merge(counts_total, counts_DMSO_R4[, c("geneID", "Counts_DMSO_R4")], by = "geneID", all = TRUE)
counts_total <- merge(counts_total, counts_DMSO_R5[, c("geneID", "Counts_DMSO_R5")], by = "geneID", all = TRUE)
counts_total <- merge(counts_total, counts_PROTAC_R3[, c("geneID", "Counts_PROTAC_R3")], by = "geneID", all = TRUE)
counts_total <- merge(counts_total, counts_PROTAC_R4[, c("geneID", "Counts_PROTAC_R4")], by = "geneID", all = TRUE)
counts_total <- merge(counts_total, counts_PROTAC_R5[, c("geneID", "Counts_PROTAC_R5")], by = "geneID", all = TRUE)
counts_total <- merge(counts_total, counts_UNC_R3[, c("geneID", "Counts_UNC_R3")], by = "geneID", all = TRUE)
counts_total <- merge(counts_total, counts_UNC_R4[, c("geneID", "Counts_UNC_R4")], by = "geneID", all = TRUE)
counts_total <- merge(counts_total, counts_UNC_R5[, c("geneID", "Counts_UNC_R5")], by = "geneID", all = TRUE)

head(counts_total)
dim(counts_total) # 78724    10


write.table(counts_total, file.path(working_dir, "RNAseq_perturbation_counts_total.txt"), sep = "\t", quote = FALSE, row.names = FALSE)


```


# DESeq2 analysis
```{r DESeq2 analysis}
### The count matrix
# counts_total contains gene counts with a "geneID" column
counts_matrix <- counts_total[, -1]  # Remove geneID column to leave only the counts
rownames(counts_matrix) <- counts_total$geneID  # Use geneID as row names
colnames(counts_matrix) <- c("DMSO_R3", "DMSO_R4", "DMSO_R5", "PROTAC_R3", "PROTAC_R4", "PROTAC_R5", "UNC_R3", "UNC_R4", "UNC_R5")

head(counts_matrix)

### The metadata
sample_conditions <- data.frame(
  condition = c("DMSO", "DMSO", "DMSO", "PROTAC", "PROTAC", "PROTAC", "UNC", "UNC", "UNC"),
  replicate = c("R3", "R4", "R5", "R3", "R4", "R5", "R3", "R4", "R5"),
  row.names = colnames(counts_matrix)  # Ensure row names match the column names in the counts matrix
)

head(sample_conditions, n=10)


### Create DESeqDataSet with condition and replicate in the design
dds <- DESeqDataSetFromMatrix(countData = counts_matrix,
                              colData = sample_conditions,
                              design = ~ condition + replicate)  # Include both condition and replicate in the design formula


### Running the DESeq2 analysis
dds <- DESeq(dds)

```

# Extract DESeq2 results for PROTAC versus DMSO
```{r Extract DESeq2 analysis results: PROTAC versus DMSO}
# Results for comparison: PROTAC vs DMSO
res_PROTAC_vs_DMSO <- results(dds, contrast = c("condition", "PROTAC", "DMSO"))
dim(res_PROTAC_vs_DMSO) # 78724     6
write.csv(res_PROTAC_vs_DMSO, file.path(working_dir, "RNAseq_PRC2_perturb_DESeq2_results_PROTAC_vs_DMSO_total_vf.csv"))

# Filter significant results (e.g., adjusted p-value < 0.05)
res_PROTAC_vs_DMSO_sig <- res_PROTAC_vs_DMSO[which(res_PROTAC_vs_DMSO$padj < 0.05), ]
nrow(res_PROTAC_vs_DMSO[which(res_PROTAC_vs_DMSO$padj > 0.5), ]) #8949
dim(res_PROTAC_vs_DMSO_sig) # 865     6

## log2FoldChange > 0.1
# How many UPregulated genes? (log2FoldChange > 0.1 and padj < 0.05)
upregulated_genes_protac0.1 <- res_PROTAC_vs_DMSO_sig[which(res_PROTAC_vs_DMSO_sig$log2FoldChange > 0.1), ]
dim(upregulated_genes_protac0.1)  # 684 UPregulated genes

# How many DOWNregulated genes (log2FoldChange < -0.1 and padj < 0.05)
downregulated_genes_protac0.1 <- res_PROTAC_vs_DMSO_sig[which(res_PROTAC_vs_DMSO_sig$log2FoldChange < -0.1), ]
dim(downregulated_genes_protac0.1)  # 181 downregulated genes

### Export results
write.csv(res_PROTAC_vs_DMSO_sig, file.path(working_dir, "RNAseq_PRC2_perturb_DESeq2_results_PROTAC_vs_DMSO_signif_vf.csv"))
write.csv(upregulated_genes_protac0.1, file.path(working_dir, "upregulated_genes_log2FC0.1_PROTAC_vs_DMSO_vf.csv"))
write.csv(downregulated_genes_protac0.1, file.path(working_dir, "downregulated_genes_log2FC0.1_PROTAC_vs_DMSO_vf.csv"))

```


# Extract DESeq2 results for UNC1999 versus DMSO
```{r Extract DESeq2 analysis results: UNC versus DMSO}

# Results for comparison: UNC vs DMSO
res_UNC_vs_DMSO <- results(dds, contrast = c("condition", "UNC", "DMSO"))
dim(res_UNC_vs_DMSO) # 78724     6
write.csv(res_UNC_vs_DMSO, file.path(working_dir, "RNAseq_PRC2_perturb_DESeq2_results_UNC1999_vs_DMSO_total_vf.csv"))

# Filter significant results (e.g., adjusted p-value < 0.05)
res_UNC_vs_DMSO_sig <- res_UNC_vs_DMSO[which(res_UNC_vs_DMSO$padj < 0.05), ]
nrow(res_UNC_vs_DMSO[which(res_UNC_vs_DMSO$padj > 0.5), ]) #8178
dim(res_UNC_vs_DMSO_sig) # 2859     6

## log2FoldChange > 0.1
# How many UPregulated genes? (log2FoldChange > 0.1 and padj < 0.05)
upregulated_genes_unc0.1 <- res_UNC_vs_DMSO_sig[which(res_UNC_vs_DMSO_sig$log2FoldChange > 0.1), ]
dim(upregulated_genes_unc0.1)  # 1783 UPregulated genes

# How many DOWNregulated genes (log2FoldChange < -0.1 and padj < 0.05)
downregulated_genes_unc0.1 <- res_UNC_vs_DMSO_sig[which(res_UNC_vs_DMSO_sig$log2FoldChange < -0.1), ]
dim(downregulated_genes_unc0.1)  # 1076 downregulated genes

### Export results
write.csv(res_UNC_vs_DMSO_sig, file.path(working_dir, "RNAseq_PRC2_perturb_DESeq2_results_UNC_vs_DMSO_signif_vf.csv"))
write.csv(upregulated_genes_unc0.1, file.path(working_dir, "upregulated_genes_log2FC0.1_UNC_vs_DMSO_vf.csv"))
write.csv(downregulated_genes_unc0.1, file.path(working_dir, "downregulated_genes_log2FC0.1_UNC_vs_DMSO_vf.csv"))

```


# Visualisation: Number of significantly differentially expressed genes (|log₂FC| ≥ 0.1, adjusted p-value < 0.05) measured by RNA-seq after PRC2 PROTAC or PRC2 catalytic inhibitor treatment relative to DMSO control. 

Light grey bars: significantly downregulated genes; dark grey bars: significantly upregulated genes.
```{r Generate a barplot}

## Summarise DEG counts in a data frame
# PROTAC perturbation
deg_counts_PROTAC <- data.frame(
  Threshold = c("log2FC > 0.1", "log2FC > 0.1"),
  Regulation = c("Up", "Down"),
  Count = c(
    nrow(upregulated_genes_protac0.1),
    nrow(downregulated_genes_protac0.1)
  )
)

deg_counts_PROTAC$Condition <- "PRC2 PROTAC"

# Catalytic inhibitor perturbation (UNC1999)
deg_counts_UNC <- data.frame(
  Threshold = c("log2FC > 0.1", "log2FC > 0.1"),
  Regulation = c("Up", "Down"),
  Count = c(
    nrow(upregulated_genes_unc0.1),
    nrow(downregulated_genes_unc0.1)
  )
)

deg_counts_UNC$Condition <- "PRC2 inhibitor"

## Combine
deg_counts_all <- rbind(deg_counts_PROTAC, deg_counts_UNC)
deg_counts_all$Condition <- factor(
  deg_counts_all$Condition,
  levels = c("PRC2 PROTAC", "PRC2 inhibitor")
)

## Barplot
pdf(file.path(working_dir, "SuppFigureS3D.pdf"), width = 5, height = 5)
ggplot(deg_counts_all, aes(x = Condition, y = Count, fill = Regulation)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("Up" = "darkgrey", "Down" = "lightgrey")) +
  theme_minimal() +
  labs(
    title = "DEG counts at log2FC > 0.1 threshold",
    x = "Condition",
    y = "Number of Significant Genes",
    fill = "Regulation"
  )
dev.off()


```


# Volcano plots of differential gene expression after PRC2 PROTAC or PRC2 catalytic inhibitor treatment, compared to DMSO. 
Blue dots indicate significantly downregulated genes (padj < 0.05 & log₂FC ≤ -0.1); orange dots indicate significantly upregulated genes (padj < 0.05 & log₂FC ≥ 0.1); the rest of the genes are shown as grey dots. Dashed grey lines indicate significance and fold-change thresholds. Labels are shown for genes with padj < 1e−3 and |log₂FC| > 0.41, corresponding to a ~30% change in expression.

```{r Volcano plot}

## Prepare Ensembl IDs
ensembl_ids_PROTAC <- rownames(res_PROTAC_vs_DMSO)
ensembl_ids_UNC <- rownames(res_UNC_vs_DMSO)

# Remove version suffixes if present (e.g., ENSG00000141510.15 -> ENSG00000141510)
ensembl_ids_clean_PROTAC <- sub("\\..*", "", ensembl_ids_PROTAC)
ensembl_ids_clean_UNC <- sub("\\..*", "", ensembl_ids_UNC)

## Map Ensembl IDs to HGNC gene symbols
gene_symbols_PROTAC <- mapIds(
    org.Hs.eg.db,
    keys = ensembl_ids_clean_PROTAC,
    column = "SYMBOL",
    keytype = "ENSEMBL",
    multiVals = function(x) x[1]  # pick first if multiple mappings
)

gene_symbols_UNC <- mapIds(
    org.Hs.eg.db,
    keys = ensembl_ids_clean_UNC,
    column = "SYMBOL",
    keytype = "ENSEMBL",
    multiVals = function(x) x[1]  # pick first if multiple mappings
)

## Add gene symbol column
res_with_symbols_PROTAC <- as.data.frame(res_PROTAC_vs_DMSO)
res_with_symbols_PROTAC$symbol <- gene_symbols_PROTAC[ensembl_ids_clean_PROTAC]

res_with_symbols_UNC <- as.data.frame(res_UNC_vs_DMSO)
res_with_symbols_UNC$symbol <- gene_symbols_UNC[ensembl_ids_clean_UNC]


# 3. Add category column for coloring points
res_with_symbols_PROTAC$category <- ifelse(
  res_with_symbols_PROTAC$padj < 0.05 & res_with_symbols_PROTAC$log2FoldChange >= 0.1, "Upregulated",
  ifelse(res_with_symbols_PROTAC$padj < 0.05 & res_with_symbols_PROTAC$log2FoldChange <= -0.1, "Downregulated", "Not significant")
)

res_with_symbols_UNC$category <- ifelse(
  res_with_symbols_UNC$padj < 0.05 & res_with_symbols_UNC$log2FoldChange >= 0.1, "Upregulated",
  ifelse(res_with_symbols_UNC$padj < 0.05 & res_with_symbols_UNC$log2FoldChange <= -0.1, "Downregulated", "Not significant")
)


## Optional: remove labels that are just Ensembl IDs
res_with_symbols_PROTAC$label <- ifelse(
  is.na(res_with_symbols_PROTAC$symbol), 
  NA, 
  res_with_symbols_PROTAC$symbol
)

res_with_symbols_UNC$label <- ifelse(
  is.na(res_with_symbols_UNC$symbol), 
  NA, 
  res_with_symbols_UNC$symbol
)

## Filter out NA padj values once, so they don't show as vertical lines at 0 on -log10(padj)
res_with_symbols_PROTAC <- subset(res_with_symbols_PROTAC, !is.na(padj))
res_with_symbols_UNC <- subset(res_with_symbols_UNC, !is.na(padj))

## Define the data frame of genes to label (significant hits with large effect size)
label_data_PROTAC <- subset(
  res_with_symbols_PROTAC,
  padj < 1e-3 & abs(log2FoldChange) > 0.41 & !is.na(label)
)

label_data_UNC <- subset(
  res_with_symbols_UNC,
  padj < 1e-3 & abs(log2FoldChange) > 0.41 & !is.na(label)
)

### Plot volcano with gene symbols as text labels for significant genes

## PROTAC
pdf(file.path(working_dir, "Volcanoplot_PROTACvsDMSO_vf.pdf"), width = 7, height = 7)
ggplot(res_with_symbols_PROTAC, aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(aes(color = category), size = 1.2, alpha = 0.6) +
  
# Custom colors
  scale_color_manual(values = c("Upregulated" = "#D55E00", "Downregulated" = "#0072B2", "Not significant" = "grey70")) +
  
# Add vertical and horizontal thresholds for padj and FC cutoffs:
  geom_vline(xintercept = c(-0.1, 0.1), linetype = "dashed", color = "black", size = 0.3) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black", size = 0.3) +
  
# Labels for selected significant genes
  geom_text(
    data = label_data_PROTAC,
    aes(label = label),
    size = 3,
    vjust = -0.5,
    check_overlap = TRUE
  ) +
  
# Axis limits and breaks
  scale_x_continuous(limits = c(-2, 2)) +
  scale_y_continuous(limits = c(0, 20), breaks = seq(0, 20, by = 4)) +

# Clean theme
  theme_minimal(base_size = 10) +
  theme(
    legend.position = "bottom",
    panel.grid.major = element_line(size = 0.2, color = "grey90"),
    panel.grid.minor = element_blank(),
    axis.line = element_line(size = 0.3, color = "black"),
    axis.ticks = element_line(size = 0.3, color = "black")
  ) +
  
  labs(
    title = "PROTAC vs DMSO: Differential Expression",
    x = expression(log[2]*" Fold Change"),
    y = expression(-log[10]*" Adjusted P-value"),
    color = "Categories"
  )
dev.off()


## UNC1999
pdf(file.path(working_dir, "Volcanoplot_UNC1999vsDMSO_vf.pdf"), width = 7, height = 7)
ggplot(res_with_symbols_UNC, aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(aes(color = category), size = 1.2, alpha = 0.6) +
  
# Custom colors
  scale_color_manual(values = c("Upregulated" = "#D55E00", "Downregulated" = "#0072B2", "Not significant" = "grey70")) +
  
# Add vertical and horizontal thresholds for padj and FC cutoffs:
  geom_vline(xintercept = c(-0.1, 0.1), linetype = "dashed", color = "black", size = 0.3) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black", size = 0.3) +
  
# Labels for selected significant genes
  geom_text(
    data = label_data_UNC,
    aes(label = label),
    size = 3,
    vjust = -0.5,
    check_overlap = TRUE
  ) +
  
# Axis limits and breaks
  scale_x_continuous(limits = c(-2, 2)) +
  scale_y_continuous(limits = c(0, 20), breaks = seq(0, 20, by = 4)) +

# Clean theme
  theme_minimal(base_size = 10) +
  theme(
    legend.position = "bottom",
    panel.grid.major = element_line(size = 0.2, color = "grey90"),
    panel.grid.minor = element_blank(),
    axis.line = element_line(size = 0.3, color = "black"),
    axis.ticks = element_line(size = 0.3, color = "black")
  ) +
  
  labs(
    title = "Catalytic inhibitor vs DMSO: Differential Expression",
    x = expression(log[2]*" Fold Change"),
    y = expression(-log[10]*" Adjusted P-value"),
    color = "Categories"
  )

dev.off()

```

